A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-09-18 manuscript/paper.Rmd
update paper

<!-- A common trajectory analysis framework -->
Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`). In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values. In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.

→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors. This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below). To wrap each method, we based ourselves on example scripts or tutorials which are mentioned in the comments of each wrapper script. We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. These discussion can be viewed on github (`r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`). About ⅓ of authors responded, and we improved the wrappers based on their feedback. 
←
### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information, though this is not required. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

>>>>

<!-- A common trajectory analysis framework -->
Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017] or module detection [@saelens_comprehensiveevaluationmodule_2018]. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`). In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values. In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.

→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below). We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. About ⅓ of authors responded, and we improved the wrappers based on their feedback. These discussion can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`. 
←
### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information, though this is not required. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
